---
title: "Tree Tomography Data Work"
author: "Max Lutz"
date: "2025-09-04"
output: html_document
---

###Soil Moisture

```{r}
library(tidyverse)
#Reading in the data from the soil flux measurements. This data was previously collected with bad chamber volume but was recalculated to correct for the error. However, we are mostly concered with the soil moisture anyway.

soil_moisture <- read.csv("~/Documents/Harvard_Forest/Data/Full_datasets/volume_recalced_soil_flux.csv") %>% mutate(., DateTime = as.POSIXct(DATE_TIME.initial_value, format = "%m/%d/%y %H:%M")) %>% 
  select(LABEL, DateTime, SWC_2.mean, EC_2.mean) %>% #selecting the columns I want (excluding soil flux)
  filter(nchar(LABEL) == 3) %>% #getting rid of some bad Labels
  subset(., .$SWC_2.mean >= 0) #Real measurements cannot be greater than 0 (ex: -9999 were present)

#Sub-setting the soil moisture to a single year span (July 2nd, 2024 - July 2nd, 2025) to give even seasonal weight.
date_range_start <- as.POSIXct("2024-07-02 00:00:00")
date_range_end <- as.POSIXct("2025-07-02 23:59:59")

soil_moisture <- soil_moisture %>% 
  subset(., (.$DateTime >= date_range_start) & (.$DateTime <= date_range_end))




#this value is the mean of soil moisture values from Stevens HydraProbe in (m3/m3) for each observation.
soil_moisture_mean <- soil_moisture %>%
  group_by(LABEL) %>%
  summarise(
    mean_Soil_Water_Content_m3pm3 = mean(SWC_2.mean, na.rm = TRUE),
    sd_Soil_Water_Content_m3pm3   = sd(SWC_2.mean, na.rm = TRUE),
    mean_Soil_Electrical_Conductivity_Spm = mean(EC_2.mean, na.rm = TRUE),
    sd_Soil_Electrical_Conductivity_Spm = sd(EC_2.mean, na.rm = TRUE)
  )

#There are be some trees in this dataset which were only visited once, and are not represented in the Tomography data set

``` 

##DBH and Species information

Now Lets add the tree demographic data
```{r}
tree_ID <- read.csv("~/Documents/Harvard_Forest/Tree_Tomography/Flux_Trees_ID.csv") %>% 
  .[c(1:60),c(1:5)]

```

#Tree Stem Flux (new 7810)

This is just for fun and personal experiment. Previous correlations did not work from the previous instrument, just want to check with 7810 data

```{r}
stem_flux <- read.csv("~/Documents/Harvard_Forest/R/Tree_Fluxes/all_7810_tree_flux_combined.csv")

stem_flux_mean <- stem_flux %>%
  group_by(REMARK) %>%
  summarise(
    stemflux_C02_mean = mean(CO2_flux_umolpm2ps, na.rm = TRUE),
    stemflux_CH4_mean = mean(CH4_flux_nmolpm2ps, na.rm = TRUE),
    stemflux_CO2_sd = sd(CO2_flux_umolpm2ps, na.rm = TRUE),
    stemflux_CH4_sd = sd(CH4_flux_nmolpm2ps, na.rm = TRUE)
  )

```


#Combinding into a dataset

```{r}

tomograph_tree_list <- read.csv("~/Documents/Harvard_Forest/Tree_Tomography/Tomography_categories_Grace.xlsx - Sheet1.csv") %>% #This is to insure that the trees measured are retained throughout the data merge.
select(Tree.ID, Density)

tomograph_tree_list$Tree.ID <- as.character(tomograph_tree_list$Tree.ID)

#Setting naming convention
soil_moisture_mean <- soil_moisture_mean %>% 
  rename(
    Tree.ID = LABEL
  ) 

soil_moisture_mean$Tree.ID <- as.character(soil_moisture_mean$Tree.ID)

#setting naming convention
tree_ID <- tree_ID %>% 
  rename(
    Tree.ID = Tag
  )

tree_ID$Tree.ID <- as.character(tree_ID$Tree.ID)

#setting naming convention
stem_flux_mean <- stem_flux_mean %>% 
  rename(
    Tree.ID = REMARK
  )

stem_flux_mean$Tree.ID <- as.character(stem_flux_mean$Tree.ID)

#Merging the data sets based on Tree.ID

tomdat_df <- left_join(tomograph_tree_list, soil_moisture_mean, by = "Tree.ID") %>% 
  left_join(., tree_ID, by = "Tree.ID") %>% 
  left_join(., stem_flux_mean, by = "Tree.ID")

#removing unwanted columns and re-aranging the columns
tomdat_df <- tomdat_df %>% 
  select(-c(Old.tag, Density)) %>%
  select(Tree.ID, species, dbh, Plot, everything())


#Exporting as a csv

write.csv(tomdat_df, file = "~/Documents/Harvard_Forest/Tree_Tomography/Tomography_suplimentary_data.csv")


```


----

##Actual Analysis and data exploration time

Binning the percent decay class by soil moisture

```{r}
library(ROCR)
library(TOC)

#Processed ERT images
absolute_ERT <- read.csv("~/Documents/Harvard_Forest/Tree_Tomography/Image Processing/Absolute_ERT_results_2025-10-10.csv") %>% 
  rename(Tree.ID = tree,
         mean_ert = mean,
         median_ert = median,
         sd_ert = sd,
         cv_ert = cv,
         gini_ert = gini,
         cma_ert = cma
  )

#soil moisture, soil conductivity, and mean stem flux
supplimentary_dat <- read.csv("~/Documents/Harvard_Forest/Tree_Tomography/Tomography_suplimentary_data.csv")

#presences absence data of
tomography_categories <- read.csv("~/Documents/Harvard_Forest/Tree_Tomography/Tomography_categories_Grace.xlsx - Sheet1.csv")


#Experimentation with TOC curve
#Need to combine data frames into a single sheet to input into the TOC generator

combo <- left_join(supplimentary_dat, tomography_categories, "Tree.ID") %>% 
  mutate(Absolute_boolean = as.numeric(recode(Moisture_Absolute, "Present" = 1, "Absent" = 0))) %>% 
  left_join(., absolute_ERT)
  


tocd <- TOC(combo$mean_Soil_Water_Content_m3pm3, combo$Absolute_boolean)


plot(tocd)

#Testing by Species

sub_combo <- subset(combo, combo$species == "bg")

tocd_sub <- TOC(sub_combo$mean_Soil_Water_Content_m3pm3, sub_combo$Absolute_boolean)

plot(tocd_sub)

```




-----


Visualization creation from meeting on 12/5/2025


% decay and moisture plot






```{r}
library(tidyverse)

Absolute_ert_moisture <- read.csv("~/Documents/Harvard_Forest/Tree_Tomography/Image Processing/Absolute_ERT_results_2025-10-10.csv") %>% 
  select(-c(filename,))

Tree_damage <- read.csv("~/Documents/Harvard_Forest/Tree_Tomography/Tree_percent_damage_Grace.csv") %>% 
  select(-notes)

tom_df <- dplyr::left_join(Absolute_ert_moisture, Tree_damage)


#Using Marra et al., 2018 as a reference, lets make some decay classes
#No Decay = Sound Wood(no rot) & No Moisture
#Incipient Decay = Sound Wood(no rot) & Moisture
#Active Decay = rot present & Moisture
#Cavity = rot presnet & no Moisture

tom_df <-  tom_df %>% 
  mutate(decay_class = NA)

for(i in 1:nrow(tom_df)){
  if(tom_df$presence.absence.rot[i] == "absent" && tom_df$moisture_absolute[i] == "absent"){
    tom_df$decay_class[i] <- "No Decay"
  } else if(tom_df$presence.absence.rot[i] == "absent" && tom_df$moisture_absolute[i] == "present"){
    tom_df$decay_class[i] <- "Incipient Decay"
  } else if(tom_df$presence.absence.rot[i] == "present" && tom_df$moisture_absolute[i] == "present"){
    tom_df$decay_class[i] <- "Active Decay"
  } else if(tom_df$presence.absence.rot[i] == "present" && tom_df$moisture_absolute[i] == "absent"){
    tom_df$decay_class[i] <- "Cavity"
  } else {
    tom_df$decay_class[i] <- NA
  }
}

tom_df_graphic <- tom_df %>% 
  filter(!is.na(decay_class))


test <- tom_df_graphic %>%
  filter(plot != "BGS") %>% 
  group_by(decay_class) %>%
  summarise(n = n())


```



Explanation of variables

Basic Statistics
Mean: Average resistivity value within the tree boundary. Higher values indicate drier conditions.
Median: Middle value when all pixels are sorted. Less sensitive to outliers than mean.
SD (Standard Deviation): Measure of variability in resistivity values. Higher SD indicates more heterogeneous moisture distribution.

Heterogeneity Metrics
CV (Coefficient of Variation): SD divided by mean. Normalized measure of relative variability (0 = uniform, higher = more variable). Values typically range from 0.1 to 1.0.
Gini Coefficient: Measure of inequality in the distribution (0 = perfect equality, 1 = maximum inequality). Higher values indicate concentrated moisture in specific areas.
Entropy: Information-theoretic measure of disorder. Higher entropy indicates more random/dispersed moisture patterns.

Spatial Metrics
CMA (Central Moisture Accumulation): Proportion of the wettest pixels (lowest 30% resistivity) that are located in the central third of the tree. Values > 0.33 indicate moisture concentration in the center.

Radial Gradient: Difference between edge mean and center mean resistivity. Positive values indicate drier edges (healthy pattern), negative values indicate drier center (possible heartwood or decay).
Radial Profile: Mean resistivity values in 8 concentric rings from center to edge. Visualizes the radial moisture distribution pattern.

Quality Metrics
NPixels: Number of pixels analyzed within the tree boundary. Indicates image resolution and analysis coverage.
Interpreting Results
Healthy Trees Typically Show:
Moderate mean resistivity values (depends on species and season)
Lower CV and Gini values (more uniform moisture)
Positive radial gradient (wetter sapwood, drier heartwood)
Smooth radial profile increasing from center to edge
Signs of Potential Issues:
Very high CV or Gini (> 0.7): Indicates patchy moisture, possible decay pockets
High CMA (> 0.5): Moisture accumulation in center, possible hollowing
Negative radial gradient: Dry center, could indicate advanced heartwood decay
Irregular radial profile: Non-uniform moisture distribution, structural issues




```{r}
#Visualization


ggplot(tom_df_graphic, aes(x = percent_solid_wood, y = mean, color = decay_class)) +
  geom_point() +
  scale_color_discrete(na.translate = FALSE) +
  stat_ellipse(type = "t")+
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  scale_color_brewer(palette = "Set1") +
  # scale_x_datetime(date_breaks = "3 months", date_labels = "%b %y") +
  labs(
    x = "Percent Solid Wood",
    y = "Mean resistivity (Ω*m)",
    color = "Decay Class"
    # title = "Decay Classes",
    # subtitle = "Optional subtitle"
  ) 


```

Lets try looking at in a different kind of way


```{r}

centroids <- tom_df_graphic %>%
  group_by(decay_class) %>%
  summarise(
    x = mean(percent_solid_wood),
    y = mean(mean),
    .groups = "drop"
  )




ggplot(tom_df_graphic, aes(x = percent_solid_wood, y = mean, color = decay_class, fill = decay_class)) +
  geom_point() +
  scale_color_discrete(na.translate = FALSE) +
  stat_ellipse(type = "t",
               geom = "polygon",
    alpha = 0.2,
    linewidth = 1,
    show.legend = FALSE,
    level = 0.68,
    )+
  geom_point(
    data = centroids,
    aes(x, y),
    size = 4,
    shape = 13,
    stroke = 1.2,
    show.legend = FALSE
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  scale_color_brewer(palette = "Set1", na.translate = FALSE) +
  scale_fill_brewer(palette = "Set1") +
  # scale_x_datetime(date_breaks = "3 months", date_labels = "%b %y") +
  guides(fill = "none") + 
 scale_x_continuous(
  limits = c(0, 105),
  breaks = seq(0, 100, by = 20)
) +
  labs(
    x = "Percent Solid Wood",
    y = "Mean resistivity (Ω*m)",
    color = "Decay Class"
    # title = "Decay Classes",
    # subtitle = "Optional subtitle"
  ) 

```



###Jon Revision

I really wish I had my notes, because I can't remember what I was tasked with doing...

1. Given the overall distribution of data, just a Z-score for the overall moisture of the species.


2.For each species, percent difference from the Maximum ERT value from true healthy (600 Ohm *m)



```{r}

write.csv(tom_df, file = "~/Documents/Harvard_Forest/Tree_Tomography/temp.csv")


max_ert <- tom_df %>%
  group_by(species) %>%
  summarise(
    max_mean = max(mean, na.rm = TRUE),
    percent_diff_to_600 = (600 - max_mean) / max_mean * 100,
    .groups = "drop"
  )

min_ert <- tom_df %>%
  group_by(species) %>%
  summarise(
    min_mean = min(mean, na.rm = TRUE),
    percent_diff_to_600 = (min_mean-600) / min_mean * 100,
    .groups = "drop"
  )



df_z <- tom_df %>%
  group_by(species) %>%
  mutate(
    mean_z = (mean - mean(mean, na.rm = TRUE)) /
             sd(mean, na.rm = TRUE)
  ) %>%
  ungroup()

df_z



ggplot(df_z, aes(x = mean_z, fill = species)) +
  geom_histogram(bins = 7, color = "black", linewidth = 0.3) +
  facet_wrap(~ species, nrow = 2, ncol = 2) +
   scale_fill_brewer(palette = "Set3") +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
  
  labs(
    x = "z-score of resistivity (Ω·m) per species"
  )

```

----

The Plan:

using the tom_df, try to find the healthiest tree of each species 90% percentile of resistance (mean)


```{r}
library(tidyverse)

tom_df_2 <- tom_df %>%
  group_by(species) %>%
  mutate(abs_dif_from_healthiest_moist = max(mean) - mean) %>%
  mutate(percent_dif_healthiest_moist = round((mean / max(mean))*100, digits = 2)) %>% 
  ungroup()



ggplot(tom_df_2, aes(x = percent_dif_healthiest_moist, y = percent_solid_wood, color = decay_class)) +
  facet_wrap(vars(species)) +
  geom_point() +
  # scale_color_discrete(na.translate = FALSE) +
  # stat_ellipse(type = "t",
    #            geom = "polygon",
    # alpha = 0.2,
    # linewidth = 1,
    # show.legend = FALSE,
    # level = 0.68,
    # )+
  # # geom_point(
  #   data = centroids,
  #   aes(x, y),
  #   size = 4,
  #   shape = 13,
  #   stroke = 1.2,
  #   show.legend = FALSE
  # ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid.major = element_line(color = "grey85", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "right"
  ) +
 #  scale_color_brewer(palette = "Set1", na.translate = FALSE) +
 #  scale_fill_brewer(palette = "Set1") +
 #  # scale_x_datetime(date_breaks = "3 months", date_labels = "%b %y") +
 #  guides(fill = "none") +
 # scale_x_continuous(
#   limits = c(0, 105),
#   breaks = seq(0, 100, by = 20)
# ) +
  labs(
    x = "Percent from healthiest indevidual in species",
    y = "Percent Solid Wood"#,
   # color = "Decay Class"
    # title = "Decay Classes",
    # subtitle = "Optional subtitle"
  )


# ggplot(tom_df_2, aes(x = cma, y = mean,
#                      color = ifelse(cma > 0.33, "blue", "brown"))) +
#   geom_point()

  
#K-mean clustering . . .
  

write.csv(tom_df_2, "~/Documents/Harvard_Forest/Tree_Tomography/Image Processing/temp.csv")




```




kmean clustering

```{r}
cluster_data <- tom_df_2 %>%
  select(tree,mean, median, sd, cv, gini, entropy, cma, radialgradiant, percent_solid_wood, percent_damaged) %>% 
  drop_na()

cluster_scaled <- cluster_data %>% 
  select(-tree) %>% 
  scale()
#setting seed due to randomness in kmean for reproducability
set.seed(278613)

kmean <- kmeans(cluster_scaled, centers = 4, nstart = 100)

cluster_data$cluster <- factor(kmean$cluster)

tom_df_2$cluster <- factor(kmean$cluster)

ggplot(tom_df_2, aes(cma, gini, color = cluster)) +
  geom_point(size = 3)

```


